---
title: "Interactive multi-channel image visualization in R"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('cytoviewer')`"
author:
- name: Lasse Meyer
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: lasse.meyer@dqbm.uzh.ch
- name: Nils Eling
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: nils.eling@dqbm.uzh.ch
output:
    BiocStyle::html_document:
        toc_float: yes
bibliography: library.bib
abstract: |
    This R package supports interactive visualization of multi-channel images 
    and segmentation masks generated by imaging mass cytometry and other highly 
    multiplexed imaging techniques using shiny. The cytoviewer interface is divided
    into image-level (Composite and Channels) and cell-level visualization (Masks). 
    It allows users to overlay individual images with segmentation masks, 
    integrates well with SingleCellExperiment and SpatialExperiment objects for 
    metadata visualization and supports image downloads.
vignette: |
    %\VignetteIndexEntry{"Interactive multi-channel image visualization in R"}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE,
                        fig.retina = 0.75, crop = NULL)
library(BiocStyle)
```

```{r library, echo=FALSE}
library(cytoviewer)
```

# Introduction

This vignette introduces the `cytoviewer` package for interactive
multi-channel image visualization. Images as well as corresponding
segmentation masks generated by imaging mass cytometry (IMC) and other
highly multiplexed imaging techniques can be interactively visualized
and explored.

The `cytoviewer` package builds on top of the `r Biocpkg("cytomapper")`
Bioconductor package and extends the static visualization strategies
provided by `cytomapper` via an **interactive Shiny application**. The
`cytoviewer` package leverages the image handling, analysis and
visualization strategies provided by the `r Biocpkg("EBImage")`
Bioconductor package and offers interactive image visualization
strategies in a similar fashion as `r Biocpkg("iSEE")` for single-cell
data. In addition, building up on `r Biocpkg("SingleCellExperiment")`,
`r Biocpkg("SpatialExperiment")` and `cytomapper::CytoImageList`
classes, the `cytoviewer` package integrates into the Bioconductor
framework for single-cell and image analysis.

## Highly multiplexed imaging

Highly multiplexed imaging allows simultaneous spatially and single-cell
resolved detection of dozens of biological molecules (e.g. proteins) in
their native tissue context. As a result, these technologies allow an
in-depth analysis of complex systems and diseases such as the tumor
microenvironment **[add citation]** and type 1 diabetes progression
**[add citation]**.

Imaging-based spatial proteomics methods **[add citation]** can be
broadly divided into fluorescent cyclic approaches such as tissue-based
cyclic immunofluorescence (t-CyCIF) [@Lin2018] and one-step mass-tag
based approaches such as multiplexed ion beam imaging (MIBI)
[@Angelo2014] and IMC [@Giesen2014].

Of note, the instructions below will focus on the visualization and
exploration of **IMC data** as an example. However, data from other
technologies such as t-CyCIF or MIBI, which produce pixel-level
intensities and (optionally) segmentation masks, can be interactively
visualized with `cytoviewer` as long as they have the appropriate input
format (see Section [Data input format](#DataInputFormat)).

### Imaging mass cytometry

IMC, an advancement of CyTOF, combines antibodies tagged with
isotopically pure rare earth metals with laser ablation and
mass-spectrometry-based detection to produce high-dimensional images
[@Giesen2014]. It captures the spatial expression of over 40 proteins in
parallel at a sub-cellular resolution of 1 μm. Thus, IMC is able to
detect cytoplasmic and nuclear localization of proteins.

## Highly multiplexed image analysis

To fully leverage the information contained in IMC and multiplexed
imaging data in general, computational tools are of key importance.

The main analysis steps, irrespective of the biological question,
include 1) Visual inspection of images for quality control, 2) Image
pre-processing and segmentation and 3) Single-cell and spatial analysis
**[add citation]**.

A comprehensive end-to-end workflow for multiplexed image processing and
analysis **[cite]** with detailed information for every analysis step
can be found [here]()

Importantly, the `cytoviewer` package can support, simplify and improve
any of these analysis steps with its easy-to-use interactive
visualization interface in R.

Below we will showcase an [example workflow](#ExampleWorkflow) that
highlights the different functionalities and potential application
fields of `cytoviewer`.

## Application overview

The `cytoviewer` interface is broadly divided into
[image-level](#ImageLevel) (Composite and Channels) and
[cell-level](#CellLevel) visualization (Masks). It allows users to
overlay individual images with segmentation masks, integrates well with
`SingleCellExperiment` and `SpatialExperiment` objects for metadata
visualization and supports [image downloads](#Download).

### Data input format {#DataInputFormat}

The `cytoviewer` package combines objects of the
`r Biocpkg("SingleCellExperiment")`, `r Biocpkg("SpatialExperiment")`
and `cytomapper::CytoImageList` classes (from `r Biocpkg("cytomapper")`)
to visualize image- and cell-level information.

The `cytoviewer` function takes up to **five arguments**.

Firstly, `image` refers to a `CytoImageList` object containing one or
multiple multi-channel images where each channel represents the
pixel-intensities of one marker (proteins in IMC).

Secondly, `mask` refers to a `CytoImageList` object containing one or
multiple segmentation masks. Segmentation masks are defined as
one-channel images containing integer values, which represent the cell
ids or background.

Thirdly, the `object` entry refers to a `SingleCellExperiment` or
`SpatialExperiment` class object that cell-specific metadata in the
`colData` slot.

Lastly, to match information between the `CytoImageList` objects and the
`SingleCellExperiment`/`SpatialExperiment` object, two additional spots
can be specified:

-   `img_id`: a single character indicating the `colData` (of the
    `SingleCellExperiment`/`SpatialExperiment` object) and
    `elementMetadata` (of the `CytoImageList` object) entry that
    contains the image identifiers. These image ids have to match
    between the `SingleCellExperiment`/ `SpatialExperiment` object and
    the `CytoImageList` object.

-   `cell_id`: a single character indicating the `colData` entry that
    contains the cell identifiers. These should be integer values
    corresponding to pixel-values in the segmentation masks.

### Data input variations {#DataInputVariation}

The functionality of `cytoviewer` depends on which input objects are
user-provided. Below we describe the **four use cases** in respect to
input objects and functionality.

**1. Usage of cytoviewer with images, masks and object**

The *full* functionality of cytoviewer can be leveraged when `image`,
`mask` and `object` are provided, which is the main intended use case.

This allows [image-level](#ImageLevel) visualization (Composite and
Channels), [cell-level](#CellLevel) visualization, overlaying images
with segmentation masks as well as metadata visualization.

**2. Usage of cytoviewer with images only**

If only the `image` object is specified, [image-level](#ImageLevel)
visualization (Composite and Channels) is possible.

**3. Usage of cytoviewer with images and masks**

[Image-level](#ImageLevel) visualization (Composite and Channels),
overlaying of images with segmentation masks and
[cell-level](#CellLevel) visualization is feasible when `image` and
`mask` objects are provided.

**4. Usage of cytoviewer with masks and object**

If `mask` and `object` are specified, [cell-level](#CellLevel)
visualization as well as metadata visualization is possible.


# Example workflow {#ExampleWorkflow}

## Installation 

The `cytoviewer` package can be installed from `Bioconductor` via:

```{r bioc-install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("cytoviewer")
```

The development version of `cytoviewer` can be installed from Github via:

```{r devel-install, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

remotes::install_github("BodenmillerGroup/cytoviewer")
```

To load the package in your R session, type the following:

```{r load}
library(cytoviewer)
```

## Example dataset

For visualization purposes, we will use a toy dataset provided 
by the `r Biocpkg("cytomapper")` package. 

The dataset contains 3 images of $100\mu{m}$ x $100\mu{m}$ dimensions 
with 362 segmented cells and pixel-intensities for 5 proteins: H3, CD99, PIN, 
CD8a, and CDH. It is a small subset from a Type 1 Diabetes dataset **(cite)**. 

Pixel-level intensities for all 5 markers (5 channels) are stored in the
`pancreasImages` object.

The corresponding segmentation masks are stored in the `pancreasMasks` 
object. 

All cell-specific metadata are stored in the `colData` slot of the 
corresponding `SingleCellExperiment` object: `pancreasSCE`. 

For more detailed information on the dataset, please refer to the 
respective documentation (e.g. via `?pancreasImages` or the vignette of the
`r Biocpkg("cytomapper")` package). 

```{r dataset}
# Load example datasets 
data("pancreasImages")
data("pancreasMasks")
data("pancreasSCE")

pancreasImages
pancreasMasks 
pancreasSCE
```

## Interface

## Image-level visualization {#ImageLevel}

## Cell-level visualization {#CellLevel}

## Image download {#Download}

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
