---
title: "Interactive multi-channel image visualization in R"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('cytoviewer')`"
author:
- name: Lasse Meyer
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: lasse.meyer@dqbm.uzh.ch
- name: Nils Eling
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: nils.eling@dqbm.uzh.ch
output:
    BiocStyle::html_document:
        toc_float: yes
abstract: |
    TODO
vignette: |
    %\VignetteIndexEntry{"Interactive multi-channel image visualization in R"}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE,
                        fig.retina = 0.75, crop = NULL)
library(BiocStyle)
```

```{r library, echo=FALSE}
library(cytoviewer)
```

# Introduction

TODO

The `cytoviewer` package builds on top of the `r Biocpkg("cytomapper")`
Bioconductor package and extends the static visualization strategies provided by
`cytomapper` via an interactive Shiny application. The `cytoviewer` package
leverages the image handling, analysis and visualization strategies provided by
the `r Biocpkg("EBImage")` Bioconductor package and offer interactive image
visualization strategies in a similar fashion as `r Biocpkg("iSEE")` for
single-cell data. In addition, building up on 
`r Biocpkg("SingleCellExperiment")`, `r Biocpkg("SpatialExperiment")` and
`cytomapper::CytoImageList` classes, the `cytoviewer` package integrates into
the Bioconductor framework for single-cell and image analysis.

# Installation

```{r installation, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("cytoviewer")
```

# Usage

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
